---
title: "Project 3 Collaboration Data Acquisition"
author: "Mei Qi Ng"
date: now
output: html_document
format:
  html:
    embed-resources: true
    toc: true
    toc-expand: true
    code-fold: show
  pdf:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Sources

(1) [*Kaggle Data Science Job Data set (Primary)*](https://www.kaggle.com/datasets/nikhilbhathi/data-scientist-salary-us-glassdoor) *: [data_cleaned csv](https://raw.githubusercontent.com/meiqing39/607-Project-3/refs/heads/main/data_cleaned_2021.csv)* (on Github repository)

    **Why this?**

    Large scale job posting dataset (job titles, salaries, skill flags (Python, SQL, AWS, Tableau etc.) and job descriptions - great structured columns ready to be analyzed, salary ranges/average - good for exploring skill-salary association, Job description for texting mining soft skills, Many job posting with variety, technical skills listed can be used for statistical relationship analysis

(2) BuiltIn Job Postings (Supplement Data Source): SoFi, TempusAI, Grow Therapy, Braze, PRC..etc

(3) Supplement Skill Reference Articles (Qualitative reference sources),

    a)  NYIT Online – \*Data Science Skills\* 

    <https://online.nyit.edu/blog/data-science-skills>

    provides overview of core technical and soft skills expected of data science professionals (programming, statistics, communication, and domain knowledge)

    b)  Pragmatic Institute – \*Must‑Have Data Science Skills for 2023\* 

    <https://www.pragmaticinstitute.com/resources/articles/data/must-have-data-science-skills-for-2023/>

    Highlights industry‑aligned skill requirements, modern tools, business communication, and deployment skills.

    **Why this?**

    Up to date skill requirements reflecting 2025-2026 job market, Includes qualitative details (soft skill, modern tools, expectation. Great for comparison between Kaggle dataset from the past and present day job demands on what quality is still the same

**Kaggle Data Acquistion Extraction Plan:**

load from csv on github, extract skill indicators, salary information for skill to salary analysis, Soft skill via text search (Teamwork, Communication..etc), grouping job titles (data scientist vs analyst vs engineers)

**BuiltIn Job Posting Extraction Plan:**

Scrapping –\> Tools (PowerBI, Tableu etc), Soft skills (Presentation, communication, collaboration et), Salary, Job description text, Required Programming Skills (R, Python, SQL, Scala etc), Industry domain, Job seniority

**Storing dataset in standardized columns**

job_idsource = "builtin"

company

job_title

skill_list

soft_skills

salary_range

job_description

date_collected

## Load Dataset and required packages

```{r}
# For data wrangling
#install.packages('polite')
library(tidyverse)
library(janitor)

# Website scrapping, HTML parsing (polite to avoid blockage)
library(rvest)
library(polite)

# 607 Project 3 Main Github repository


kaggle_url <- "https://raw.githubusercontent.com/meiqing39/607-Project-3/main/data_cleaned_2021.csv"

kg <- read_csv(kaggle_url, show_col_types = FALSE) |> 
  clean_names()   # <- renames by snake_case format, removes special character

#Ensures consistent and reproducible references to variable by using janitor function clean_names()
```

```{r}
# Tibble Resource table
sources_roles <- tibble(
  Source = c(
    "Kaggle dataset (structured job postings)",
    "BuiltIn job postings (recent, unstructured)",
    "NYIT article (skills overview)",
    "Pragmatic Institute article (skills 2023)"
  ),
  Role = c(
    "Primary quantitative dataset used for skill frequency and salary association analyses",
    "Supplemental contemporary postings used to validate/contrast 2025–2026 skill demand",
    "Qualitative reference to define/justify skill taxonomy (not stored in DB)",
    "Qualitative reference to contextualize/interpret findings (not stored in DB)"
  )
)

print(sources_roles)
```

## Entity-Relationship (ER) Diagram

ER diagram shows the normalized database design used to store job postings, technical skills, and soft skills collected from the Kaggle dataset and BuiltIn job postings.

```{r, echo=FALSE, out.width="90%"}

photoURL <- "https://github.com/meiqing39/607-Project-3/blob/main/Project%203%20ERD.png?raw=true"
knitr::include_graphics(photoURL)
```

## Normalize Kaggle CSV - relational connection job_id

-   job level attributes (title, company, industry, salary) extracted into jobs table

-   Technical skills: extract from binary columns, reshape into longer format with pivot longer function connect each job with required skills

-   job description test extracted without modding for later soft skill text analysis

```{r}
# Extract job level data
jobs <- kg |> 
  transmute(
    job_id = row_number(),
    job_title,
    company = company_name,
    location,
    industry,
    avg_salary_k,
    source = "kaggle"
  )

# Extract job description (soft skill use)
job_descriptions <- kg |> 
  transmute(
    job_id = row_number(),
    job_description
  )

# Id skill columns

skill_cols <- names(kg)[
  names(kg) %in% c(
    "python","sql","aws","excel","tableau","spark",
    "sas","pytorch","keras","scikit","tensor",
    "hadoop","flink","mongo","google_an"
  )
]



# Combine created skill_cols to job_id to make job_skills table
job_skills <- kg |> 
  mutate(job_id = row_number()) |> 
  select(job_id, all_of(skill_cols)) |> 
  pivot_longer(
    cols = -job_id,
    names_to = "skill",
    values_to = "present"
  ) |> 
  filter(present == 1) |> 
  select(job_id, skill)
         
```

## Job Posting Web Scrapping

```{r}

# Create Scrapping function for scrapping of job title and job description later

get1 <- function(html, css) {
  node <- html_element(html, css)
  if (inherits(node, "xml_missing") || is.na(node)) return("")
  html_text2(node)
}

first_nonempty <- function(html, selectors) {
  for (sel in selectors) {
    txt <- get1(html, sel)
    if (nzchar(txt)) return(txt)
  }
  ""
}

scrape_builtin <- function(url, ua = "607-Project-3/edu (rvest+polite)") {
  ses <- polite::bow(url, user_agent = ua, force = TRUE)
  pg  <- polite::scrape(ses)

  # Title
  title <- first_nonempty(pg, c("h1", "meta[property='og:title']"))
  if (!nzchar(title)) title <- NA_character_

  # Description: try a few common containers, fall back to main/body
  desc <- first_nonempty(pg, c(
    "article",
    "[data-testid='job-description']",
    "[data-qa='job-description']",
    "div[class*='job-description']",
    "section[class*='description']"
  ))
  if (!nzchar(desc)) desc <- get1(pg, "[role='main']")
  if (!nzchar(desc)) desc <- get1(pg, "body")
  if (!nzchar(desc)) desc <- NA_character_

  tibble(
    job_title       = title,
    job_description = desc,
    job_url         = url
  )
}

# Organize all job posting URLs into a tibble

builtin_urls <- tibble(
  job_url = c(
    "https://builtin.com/job/manager-data-scientist-alternate-data-strategy/8162509",
    "https://builtin.com/job/principal-associate-data-scientist-ai-software-engineering/8017518",
    "https://builtin.com/job/principal-associate-data-scientist-us-card-new-credit-team/8029691",
    "https://builtin.com/job/senior-associate-data-scientist/8432791",
    "https://builtin.com/job/data-scientist-ii-real-world-evidence-rwe-pharma-r-d/8031762",
    "https://builtin.com/job/staff-data-scientist-growth-care/8019471",
    "https://www.builtinnyc.com/job/senior-staff-data-scientist-invest/8697390",
    "https://www.builtinnyc.com/job/staff-data-scientist-borrow/8290699",
    "https://www.builtinnyc.com/job/field-data-scientist-ai-deployment/7405417",
    "https://www.builtinnyc.com/job/ai-genai-data-scientist-senior-manager/8258875",
    "https://www.builtinnyc.com/job/data-scientist/8446518"
  ))


# Scrape --> builtin_raw 
safe_scrape <- purrr::safely(scrape_builtin)

builtin_raw <- builtin_urls |> 
  mutate(res = map(job_url, safe_scrape)) |> 
  transmute(
    job_url,
    job_title  = map_chr(res, ~ .x$result$job_title %||% NA_character_),
    job_description = map_chr(res, ~ .x$result$job_description %||% NA_character_))


# Double checking that job description returns with page body text for skill, soft skill keywords, and frequency comparison later - text number appear = good 

nchar(builtin_raw$job_description)


# Scrape and  Manually add company and industry from data seen on job posting website
builtin_jobs <- builtin_raw |> 
  mutate(
    company = case_when(
      str_detect(job_url, "capital|cap1") ~ "Capital One",
      str_detect(job_url, "nylife|new-york-life|senior-associate-data-scientist/8432791")
~ "NY Life",
      str_detect(job_url, "tempus") ~ "Tempus AI",
      str_detect(job_url, "grow") ~ "Grow Therapy",
      str_detect(job_url, "sofi") ~ "SoFi",
      str_detect(job_url, "braze") ~ "Braze",
      str_detect(job_url, "prc") ~ "PRC",
      str_detect(job_url, "octus") ~ "Octus",
      TRUE ~ NA_character_
    ),
    industry = case_when(
      company %in% c("Capital One", "SoFi", "NY Life") ~ "Finance",
company == "Tempus AI" ~ "Healthcare / Pharma",
      company == "Grow Therapy" ~ "Healthcare",
      company == "Braze" ~ "SaaS / Martech",
      company %in% c("PRC", "Octus") ~ "Consulting",
      TRUE ~ NA_character_
    ),
    source = "builtin",
    job_id = row_number()
  ) %>%
  select(job_id, job_title, company, industry, source, job_url, job_description)


# extract skills seen on job descriptions
skill_keywords <- c(
  # languages
  "python","r","sql","scala","java",
  # libraries / ml
  "pytorch","tensorflow","scikit","sklearn","xgboost",
  # platforms
  "aws","azure","gcp","databricks","snowflake","spark",
  # bi / viz
  "tableau","power bi")


builtin_job_skills <- builtin_jobs %>%
  transmute(job_id, text = tolower(job_description %||% "")) %>%
  filter(nzchar(text)) %>%
  crossing(skill = skill_keywords) %>%
  mutate(rx = paste0("\\b", skill, "\\b")) %>%
  filter(str_detect(text, regex(rx, ignore_case = TRUE))) %>%
  distinct(job_id, skill)

```

# Table result to ensure data was extracted and organized

```{r}

head(builtin_jobs)

builtin_job_skills |>
  count(skill) |> 
  arrange(desc(n))
```

##Extracting csv file from builtin_jobs and builtin_job_skills

```{r}
write_csv(builtin_jobs, "builtin_jobs.csv")
write_csv(builtin_job_skills, "builtin_job_skills.csv")
```

I scraped job posting descriptions from Builtin to extract modern technical and soft skills using keyword matching while also understanding that some attributes (salary, explicit seniority levels) are inconsistent available on job posting websites. Salary based analyses rely mostly on the structured kaggle dataset, while the Builtin job posting provide up to date information on skill demands and tools used as data scientists. Industry classification were assigned manually based on company information.

```{r}
pacman::p_load(janitor,
               tidyverse)

#library(tidyverse) Moved into pacman statement package to make it easier to open up and install the latest version
#library(dplyr) Removed this library because it is already included in tidyverse
#library(readr) Removed this library because it is already included in tidyverse
#library(ggplot2) Removed this library because it is already included in tidyverse
#library(janitor) Moved into pacman statement package to make it easier to open up and install the latest version

kaggle_url <- "https://raw.githubusercontent.com/meiqing39/607-Project-3/refs/heads/main/data_cleaned_2021.csv"
kg <- read_csv(kaggle_url, show_col_types = FALSE) |> 
  clean_names()   # <- renames by snake_case format, removes special character

#Ensures consistent and reproducible references to variable by using janitor function clean_names()
kg

colnames(kg)
#head(10)
```

# **Data quality issues**

Several variables contain placeholder values that represent missing data rather than valid observations. For example, `Rating`, `Founded`, and `Age` include values of `-1`, while variables such as `Competitors`, `seniority_by_title`, and `Degree` use strings such as `"-1"` or `"na"` to indicate unavailable information. These values must be recoded as missing before further analysis.

## **Summary of the initial inspection**

The dataset contains 742 job postings and 42 variables. These variables include job and company descriptors, salary information, and binary indicators for technical skills such as Python, SQL, AWS, and Tableau. Initial inspection using `glimpse()` and `summary()` showed that the dataset contains both raw text variables and derived variables that are more suitable for analysis.

## **Skill demand**

The binary skill columns provide a practical proxy for skill demand in job postings. Since these variables are coded as 0 and 1, their means can be interpreted as the proportion of postings that require each skill. Preliminary results suggest that Python, Excel, and SQL are among the most frequently requested skills in the dataset.

## **Outliers and unusual values**

Exploratory summaries also suggest the presence of extreme values. For instance, company age ranges up to 277 years, and average salary ranges from 15.5 to 254 thousand dollars. While some of these values may be valid, they require further inspection through visualizations such as boxplots and histograms.

## **#Re-coding placeholder values as missing**

**Standardize Kaggle jobs**

```{r}
kaggle_jobs <- kg |>
  transmute(
    job_id = paste0("kg_", row_number()),
    source = "kaggle",
    job_title,
    company = company_name,
    location,
    industry,
    avg_salary_k,
    salary_min_k = lower_salary,
    salary_max_k = upper_salary,
    job_description,
    job_url = NA_character_
  )
kaggle_jobs
```

## Standardize BuiltIn jobs

```{r}
builtin_jobs_clean <- builtin_jobs |>
  transmute(
    job_id = paste0("bi_", row_number()),
    source = "builtin",
    job_title,
    company,
    location = NA_character_,
    industry,
    avg_salary_k = NA_real_,
    salary_min_k = NA_real_,
    salary_max_k = NA_real_,
    job_description,
    job_url
  )
builtin_jobs_clean
```

## Merge Job postings

```{r}
all_jobs <- bind_rows(kaggle_jobs, builtin_jobs_clean)
all_jobs
```

## **Missing Values**

Missing values were identified using the `is.na()` function. Depending on the variable, missing values were either removed or handled appropriately to maintain data integrit

## **Merge kaggle skills**

```{r}
skill_cols <- names(kg)[
  names(kg) %in% c(
    "python","sql","aws","excel","tableau","spark",
    "sas","pytorch","keras","scikit","tensor",
    "hadoop","flink","mongo","google_an"
  )
]

kaggle_job_skills <- kg |>
  mutate(job_id = paste0("kg_", row_number())) |>
  select(job_id, all_of(skill_cols)) |>
  pivot_longer(
    cols = -job_id,
    names_to = "skill",
    values_to = "present"
  ) |>
  filter(present == 1) |>
  select(job_id, skill)
```

#Summarizing the skills we can see that python, excel and sql are the most required skill in the job market

**#BuiltIn skills**

```{r}
skill_keywords <- c(
  "python","r","sql","scala","java",
  "pytorch","tensorflow","scikit","sklearn","xgboost",
  "aws","azure","gcp","databricks","snowflake","spark",
  "tableau","power bi"
)

builtin_job_skills_clean <- builtin_jobs |>
  mutate(job_id = paste0("bi_", row_number())) |>
  transmute(job_id, text = tolower(coalesce(job_description, ""))) |>
  filter(nzchar(text)) |>
  crossing(skill = skill_keywords) |>
  mutate(rx = paste0("\\b", skill, "\\b")) |>
  filter(str_detect(text, regex(rx, ignore_case = TRUE))) |>
  distinct(job_id, skill)
```

## Merge all skills

```{r}
all_job_skills <- bind_rows(
  kaggle_job_skills,
  builtin_job_skills_clean
)
all_job_skills

count(all_jobs, source)
count(all_job_skills, skill, sort = TRUE)

glimpse(all_jobs)
glimpse(all_job_skills)
```

checks

```{r}
count(all_jobs, source)
count(all_job_skills, skill, sort = TRUE)

glimpse(all_jobs)
glimpse(all_job_skills)
```

### Kaggle Frequency of skills in job postings

```{r}
skill_counts <- kaggle_job_skills |>
  count(skill, sort = TRUE)

ggplot(skill_counts, aes(x = reorder(skill, n), y = n, fill = skill )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Frequency of Skills in Job Postings",
    x = "Skill",
    y = "Number of postings"
  ) +
  theme(legend.position = 'none')
```

### BuiltIn frequency of skills in job postings

```{r}
skill_counts <- builtin_job_skills_clean |>
  count(skill, sort = TRUE)

ggplot(skill_counts, aes(x = reorder(skill, n), y = n, fill = skill )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Frequency of Skills in Job Postings",
    x = "Skill",
    y = "Number of postings"
  ) +
  theme(legend.position = 'none')
```

## **Number of jobs posting vs Average Salary**

```{r}
all_jobs_salary <- all_jobs |>
  filter(!is.na(avg_salary_k))

ggplot(all_jobs_salary, aes(x = avg_salary_k)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Average Salary",
    x = "Average Salary (K)",
    y = "Number of Job Postings"
  )

ggplot(all_jobs_salary, aes(y = avg_salary_k)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Average Salary",
    y = "Average Salary (K)"
  )
```

### Check salary outliers

##The Average Salary is 70k to 125k after observing the boxplot

The histogram shows that most job postings fall within the mid-range salary band, with a concentration between approximately 70K and 130K. Higher salary values are less frequent and may represent senior or specialized roles.

### **Check job title and salary**

```{r}
job_salary_summary <- all_jobs_salary |>
  
  group_by(job_title) |>
  summarise(
    mean_salary = mean(avg_salary_k, na.rm = TRUE),
    median_salary = median(avg_salary_k, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n >= 5) |>
  arrange(desc(mean_salary))
job_salary_summary |>
  ggplot(aes( x = reorder(job_title, mean_salary), y = mean_salary, fill = job_title)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Salary for Job Title",
    x = "Job Title",
    y = " Average Salary (K)"
  ) +
  theme(legend.position = 'none')

```

##To compare pay across roles, job postings were grouped by job title and summarized using average salary. Because many titles appeared only a few times, the analysis focused on titles with at least five postings. This provides a more stable comparison of salary levels across common job categories.

### Comparing Kaggle vs BuiltIn skills

```{r}
source_skill_counts <- all_jobs |>
  left_join(all_job_skills, by = "job_id") |>
  filter(!is.na(skill)) |>
  count(source, skill)

source_skill_counts |>
  ggplot(aes(x = reorder(skill, n), y = n, fill = source)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Skill Frequency by Data Source",
    x = "Skill",
    y = "Number of Job Postings"
  )
```

## Analysis

What are the most common required skills?

Let's look at the top 5 skills from the Builtin and Kaggle data sets:

```{r}
builtin_top5 <- source_skill_counts |> 
  filter(source == "builtin") |>
  arrange(desc(n)) |>
  rename("count" = n) |>
  slice_head(n = 5)

kaggle_top5 <- source_skill_counts |> 
  filter(source == "kaggle") |>
  arrange(desc(n)) |>
  rename("count" = n) |>
  slice_head(n = 5)

library(gt)


builtin_top5 |>
  select(skill, count) |>
  gt() |>
  cols_label(
    skill = "Skill",
    count = "Count"
  ) |>
  tab_header(title = "Top 5 Skills - Builtin")

builtin_top5 |> 
  ggplot(aes(x = reorder(skill, -count), y = count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Top 5 Skills (Builitin)", x = "Skill", y = "Count")


kaggle_top5 |>
  select(skill, count) |>
  gt() |>
  cols_label(
    skill = "Skill",
    count = "Count"
  ) |>
  tab_header(title = "Top 5 Skills - Kaggle")

kaggle_top5 |> 
  ggplot(aes(x = reorder(skill, -count), y = count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Top 5 Skills (Kaggle)", x = "Skill", y = "Count")
```

For the Builtin data set, Python, SQL, and R are the top software skills. Cloud skills are also in demand.

It's worth noting that this data set is much smaller and all the job posts are in the greater NYC area, and salaries will reflect that.

Kaggle shows Python, Excel, and SQL as the top 3 software skills. R is excluded from this data set, so it's not a good metric of how commonly R is a required skill. AWS cloud skills are also in demand, as is Apache Spark.

Looking at the full list, Tableau shows up quite early. In my experience, the barrier to using Tableau isn't necessarily learning to use it, but the price.

### All Skills

```{r}
#visualizing all Builtin Skills
source_skill_counts |> 
  filter(source == "builtin") |>
  arrange(desc(n)) |>
  rename("count" = n)  |>
  ggplot(aes(x = reorder(skill, -count), y = count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "All Skills (Builitin)", x = "Skill", y = "Count")


#All Kaggle skills
source_skill_counts |> 
  filter(source == "kaggle") |>
  arrange(desc(n)) |>
  rename("count" = n) |>
   ggplot(aes(x = reorder(skill, -count), y = count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "All Skills (Kaggle)", x = "Skill", y = "Count")

```

In general, the Kaggle skills start with broad software (Python, Excel, SQL, AWS), then moves into more niche Apache software. Some of these are advanced Python libraries, like Pytorch, scikit, and Keras. Tensor is a data structure.

## Salary Analysis

Above, we have salary by title. In general, data scientists and data engineers make more than analysts.

Here, we'll look at salary by state, just to see the impact of that and benchmark for later observations.

```{r}
all_jobs_state <- all_jobs_salary |>
    separate(col = location, into = c("city", "state"), sep = ",") |>
  mutate(across(c(city, state), str_trim)) |> 
  mutate(state = if_else(state == "Los Angeles", "CA", state))


#benchmarking the salaries for later observations

summary(all_jobs_state$avg_salary_k)

avg_state <- all_jobs_state |>
  filter(source == "kaggle") |>
  group_by(state) |>
  summarise(
    avg_salary = round(mean(avg_salary_k, na.rm = TRUE)),
    job_count = n()
  ) |>
  arrange(desc(avg_salary))

#top 10 states by salary

avg_state |>
  slice_head(n = 10) |>
  select(state, avg_salary, job_count) |>
  gt() |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  cols_label(
    state = "State",
    avg_salary = "Average Salary (k)",
    job_count = "Number of Jobs"
  ) |>
  tab_header(title = "Top 10 States by Average Salary")


avg_state |> 
  slice_head(n = 10) |> 
  ggplot(aes(x = reorder(state, -avg_salary), y = avg_salary, fill = state)) +
  geom_col() +
  geom_text(aes(label = paste0(avg_salary, "k")), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Top 10 States by Average Salary", x = "State", y = "Average Salary (k)")

```

### Skills and Salary

Here, we take the job skills table and join it with the KG skills table based on index. Then, we create an average salary per skill column. This is to see which skill is associated with the highest average salary.

```{r}
#making a slight tweak to this code so I can keep the index column and join
#with the kg data set for average salary

job_skills_2 <- kg |> 
  mutate(job_id = index) |> 
  select(job_id, all_of(skill_cols)) |> 
  pivot_longer(
    cols = -job_id,
    names_to = "skill",
    values_to = "present"
  ) |> 
  filter(present == 1) |> 
  select(job_id, skill)

#left join with KG to get the salary
job_skills_salary <- job_skills_2 |>
    left_join(kg |> select(index, avg_salary_k), by = c("job_id" = "index"))

#Averaging the salary
salary_skill <- job_skills_salary |>
  group_by(skill) |>
  summarise(
    avg_salary = round(mean(avg_salary_k, na.rm = TRUE)),
    job_count = n()) |>
      arrange(desc(avg_salary))


salary_skill |> 
    ggplot(aes(x = reorder(skill, -avg_salary), y = avg_salary, fill = skill)) +
    geom_col() +
    geom_text(aes(label = paste0(avg_salary, "k")), vjust = -0.5, size = 3) +
    scale_fill_viridis_d(option = "magma") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    guides(fill = "none") +
    labs(title = "Average Salary by Skill", x = "Skill", y = "Average Salary (k)")
```

In general, it seems like more niche skills are more highly valued. Scikit and Keras are Python libraries. Flink = Apache Flink. SAS ranks fifth – this data set is a little older and SAS is likely not as popular as it used to be.

### Are certain skills associated with lower salaries?

How does the average salary for each skill compare to the median salary?

```{r}
all_jobs_salary |> summarise(mean(avg_salary_k))
all_jobs_salary

#the average salary for this data set is ~101K

salary_skill |> 
  mutate("difference" = avg_salary - 101) |>
  mutate("percentage" = round(difference/101,2)) |>
  select(skill, avg_salary, difference, percentage) |>
  gt() |>
  cols_label(
    skill = "Skill",
    avg_salary = "Mean Salary (k)",
    difference = "Diff from Mean Salary (k)",
    percentage = "% Diff from Mean Salary"
  ) |>
  tab_header(title = "How Much Are Skills Worth?")
  

#A slightly dramatic bar chart
salary_skill |> 
  mutate(difference = avg_salary - 101) |>
  ggplot(aes(x = reorder(skill, -difference), y = difference, fill = skill)) +
  geom_col() +
  geom_text(aes(label = paste0(difference, "k"), 
                vjust = ifelse(difference >= 0, -0.5, 1.5)), 
            size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  expand_limits(y = c(min(salary_skill$avg_salary - 101) * 1.2, 
                      max(salary_skill$avg_salary - 101) * 1.2)) +
  labs(title = "How Much Are Skills Worth?", x = "Skill", y = "Difference from Mean Salary (k)")
```

Excel is the #2 skill, but it's assocated with a lower salary. Can we conclude that learning Excel will drop your salary down by two grand? Probably not. But it's a more accessible skill, and may be associated with less intense/specialized jobs. Tableau is also fairly accessible, as is Google Analytics.

### Skills and Title

What (if anything) does title mean? Are certain skills associated with a data scientist/analyst/engineer vs an analyst? On average, what is the pay differential between these titles?

Let's look at titles and skill counts together.

```{r}
#This data set that keeps everything and pivots skills
kg_long <- kg |>
    pivot_longer(
        cols = python:google_an,
        names_to = "skill",
        values_to = "present"
    ) |>
    filter(present == 1) |>
    select(index:age, job_title_sim:degree, skill)

kg_skills_pivot <- kg_long |>
  select(job_title_sim, skill)

#changing the title "data analitics" to just analyst, so it matches 
#(it seems like an error and it's also a typo)
kg_skills_pivot <- kg_skills_pivot |>
    mutate(job_title_sim = if_else(job_title_sim == "data analitics", "analyst",
                                   job_title_sim))

  

kg_skills_agg <- kg_skills_pivot |> 
    group_by(job_title_sim, skill) |> 
    summarise("skill_count" = n(), .groups = "drop")

#Let's look at analyst, data engineer, and data scientist

kg_skills_agg |> filter(job_title_sim == "analyst") |>
  arrange(desc(skill_count)) |>
   select(skill, skill_count) |>
  gt() |>
  cols_label(
    skill = "Skill",
    skill_count = "Count"
  ) |>
  tab_header(title = "Most Common Analyst Skills")


kg_skills_agg |> filter(job_title_sim == "data scientist") |>
  arrange(desc(skill_count)) |>
   select(skill, skill_count) |>
  gt() |>
  cols_label(
    skill = "Skill",
    skill_count = "Count"
  ) |>
  tab_header(title = "Most Common Data Scientist Skills")


kg_skills_agg |> filter(job_title_sim == "data engineer") |>
  arrange(desc(skill_count)) |>
   select(skill, skill_count) |>
  gt() |>
  cols_label(
    skill = "Skill",
    skill_count = "Count"
  ) |>
  tab_header(title = "Most Common Data Engineer Skills")

#analyst skills bar chart
kg_skills_agg |> 
  filter(job_title_sim == "analyst") |>
  ggplot(aes(x = reorder(skill, -skill_count), y = skill_count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = skill_count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Most Common Data Analyst Skills", x = "Skill", y = "Count")

#scientist
kg_skills_agg |> 
  filter(job_title_sim == "data scientist") |>
  ggplot(aes(x = reorder(skill, -skill_count), y = skill_count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = skill_count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Most Common Data Scientist Skills", x = "Skill", y = "Count")

#engineer
kg_skills_agg |> 
  filter(job_title_sim == "data engineer") |>
  ggplot(aes(x = reorder(skill, -skill_count), y = skill_count, fill = skill)) +
  geom_col() +
  geom_text(aes(label = skill_count), vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(title = "Most Common Data Engineer Skills", x = "Skill", y = "Count")

```

Analysts seem more likely to need excel, SQL, and Tableau before python, whereas engineers and scientists generally need SQL, Python, and cloud skills. Analysts need more visualization tools like Tableau, BI, and google analytics. In general, programming skills are preferred for scientists and engineers.

### What fields pay the most?

Is understanding a particular area (e.g. healthcare or finance) correlated with a higher salary?

In data science, you may need two skill sets. For example, you may need chemistry and data science skills in order to work at a pharmaceutical company.

```{r}
#all jobs salary
by_industry <- all_jobs_salary |> 
  filter(industry != "-1") |> 
  group_by(industry) |>
  summarise(
    avg_salary = round(mean(avg_salary_k, na.rm = TRUE), 0),
    job_count = n()
  ) |>
  arrange(desc(avg_salary))

#there are a lot of one-offs in here, so I am going to get rid of anything with a count of less than 2

by_industry_multi <- by_industry |>
  filter(job_count > 1)

#Let's look at the top 10 and bottom 10

by_industry_multi |>
  slice_head(n = 10) |>
  select(industry, avg_salary) |>
  gt() |>
  cols_label(
    industry = "Industry",
    avg_salary = "Average Salary (Thousands)"
  ) |>
  tab_header(title = "Top Average Salaries by Industry")

```

Unsurprisingly, 3/10 are in finance (brokerage, financial analytics, investment banking), industries that tend to pay more. Others are in technology generally, along with media.

```{r}
by_industry_multi |>
  slice_tail(n = 10) |>
  select(industry, avg_salary) |>
  gt() |>
  cols_label(
    industry = "Industry",
    avg_salary = "Average Salary (Thousands)"
  ) |>
  tab_header(title = "Bottom Average Salaries by Industry")
```

Some of these are industries where salaries tend to be on the lower side, like food & beverage and social services. Construction and architecture are adjacent fields.

Gambling is interesting – it may offer higher salaries now, given the growth of gambling apps.

### What field has the most jobs?

If we define valuable as "most opportunities":

```{r}
by_industry_multi |>
  arrange(desc(job_count)) |>
  slice_head(n = 10) |>
  select(industry, job_count) |>
  gt() |>
  cols_align(
    align = "center",
    columns = job_count
  ) |>
  cols_label(
    industry = "Industry",
    job_count = "Number of Jobs"
  ) |>
  tab_header(title = "Most Common Industries")


by_industry_multi |>
    arrange(desc(job_count)) |>
    slice_head(n = 10) |>
    ggplot(aes(x = reorder(industry, -job_count), y = job_count, fill = industry)) +
    geom_col() +
    geom_text(aes(label = job_count), vjust = -0.2) +
    scale_fill_viridis_d(option = "magma") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    guides(fill = "none") +
    labs(title = "Most Common Industries for Data Science", x = "Industry", y = "Number of Jobs")
```

According to this, having a biology background could be valuable (biotech & pharmaceuticals). There's demand in IT and adjacent fields like software and internet, along with insurance carriers (presumably to assess risk).

## Modeling

### Data Prep

```{r}
final_wide <- all_jobs %>%
  left_join(all_job_skills, by = "job_id") %>%
  mutate(has_skill = 1L) %>%
  pivot_wider(
    names_from = skill,
    values_from = has_skill,
    values_fill = list(has_skill = 0L),
    values_fn = list(has_skill = max)
  )

# Columns from the original all_jobs table (keep these)
orig_cols <- names(all_jobs)

# Skill columns are all columns in final_wide that were *not* in all_jobs
skill_cols <- setdiff(names(final_wide), orig_cols)

# Identify skill columns that are all zeros (no presence across any job)
zero_only <- vapply(final_wide[skill_cols], function(x) {
  # handle numeric/logical 0/1; coerce characters like "0"/"1" if needed
  if (is.character(x)) x <- suppressWarnings(as.numeric(x))
  sum(x, na.rm = TRUE) == 0
}, logical(1))

# Drop unnecessary columns
final_wide_trim <- final_wide %>%
  select(
    -any_of(skill_cols[zero_only]),   # drop skills with only 0
    -any_of("NA")                     # drop the NA column
  )
```

### Building LM Function

```{r}
run_univariate_lm <- function(df,
                              y = "avg_salary_k",
                              x_cols = 12:34,
                              na_action = na.exclude,
                              assign_to_global = FALSE,
                              sanitize_names = FALSE,
                              coerce_binary_chars = TRUE) {
  # Basic checks
  if (!y %in% names(df)) {
    stop(sprintf("Response column '%s' not found in data.", y))
  }
  if (any(x_cols < 1 | x_cols > ncol(df))) {
    stop("x_cols index range is out of bounds for the provided data frame.")
  }
  
  # Optionally sanitize names to syntactic R names
  if (sanitize_names) {
    old_names <- names(df)
    names(df) <- make.unique(make.names(names(df)))
    if (!y %in% names(df)) {
      stop("After sanitizing names, the response column was renamed; set y to the new name.")
    }
    message("Column names were sanitized via make.names() + make.unique().")
  }
  
  # Ensure response is numeric (lm works best this way)
  if (!is.numeric(df[[y]])) {
    suppressWarnings({
      df[[y]] <- as.numeric(df[[y]])
    })
    if (anyNA(df[[y]])) {
      warning(sprintf("Coercing '%s' to numeric introduced NAs; check your response column.", y))
    }
  }
  
  results <- list()
  
  for (i in x_cols) {
    xname <- names(df)[i]
    xvec <- df[[xname]]
    
    # Skip problematic types (list/data.frame columns)
    if (is.list(xvec) && !is.atomic(xvec)) {
      message(sprintf("Skipping '%s': predictor is a list/data-frame column.", xname))
      next
    }
    
    # Optional coercion for binary character columns
    if (coerce_binary_chars && is.character(xvec)) {
      vals <- unique(na.omit(xvec))
      if (all(vals %in% c("0", "1"))) {
        xvec_num <- as.numeric(xvec)
        df[[xname]] <- xvec_num
      }
    }
    
    # Skip non-variant predictors
    if (all(is.na(xvec)) || length(unique(na.omit(xvec))) < 2) {
      message(sprintf("Skipping '%s': predictor has <2 unique non-NA values.", xname))
      next
    }
    
    # Build formula with backticks to handle non-syntactic names safely
    f <- as.formula(sprintf("`%s` ~ `%s`", y, xname))
    
    # Fit model (try-catch for edge cases)
    fit <- try(lm(f, data = df, na.action = na_action), silent = TRUE)
    if (inherits(fit, "try-error")) {
      message(sprintf("Skipping '%s': lm() failed (%s).", xname, as.character(fit)))
      next
    }
    
    # Store with name "lm_<column_name>"
    obj_name <- paste0("lm_", xname)
    results[[obj_name]] <- fit
    
    # Print summary
    cat("\n", strrep("=", 72), "\n", sep = "")
    cat(sprintf("Model: %s\nFormula: %s\n", obj_name, deparse(f)))
    cat(strrep("-", 72), "\n", sep = "")
    print(summary(fit))
  }
  
  # Optionally assign each model to the global environment
  if (assign_to_global && length(results) > 0) {
    list2env(results, envir = .GlobalEnv)
    message("Models assigned to .GlobalEnv with names 'lm_<column_name>'.")
  }
  
  invisible(results)
}
```

### Running LM Function

```{r}
models <- run_univariate_lm(final_wide_trim)
```

### Building Multi LM Function

```{r}
run_multivar_lm <- function(df,
                            y = "avg_salary_k",
                            x_cols = 12:34,
                            na_action = na.exclude,
                            sanitize_names = FALSE,
                            coerce_binary_chars = TRUE,
                            standardize = FALSE) {
  # --- Basic checks ---
  if (!y %in% names(df)) {
    stop(sprintf("Response column '%s' not found in data.", y))
  }
  if (any(x_cols < 1 | x_cols > ncol(df))) {
    stop("x_cols index range is out of bounds for the provided data frame.")
  }
  
  # --- Optionally sanitize column names ---
  if (sanitize_names) {
    names(df) <- make.unique(make.names(names(df)))
    if (!y %in% names(df)) {
      stop("After sanitizing names, the response column was renamed; update 'y' to the new name.")
    }
    message("Column names were sanitized via make.names() + make.unique().")
  }
  
  # --- Ensure response is numeric ---
  if (!is.numeric(df[[y]])) {
    suppressWarnings({
      df[[y]] <- as.numeric(df[[y]])
    })
    if (anyNA(df[[y]])) {
      warning(sprintf("Coercing '%s' to numeric introduced NAs; check your response column.", y))
    }
  }
  
  # --- Collect candidate predictors ---
  xnames <- names(df)[x_cols]
  
  # --- Filter: skip problematic types, coerce binary chars, skip constants ---
  keep <- logical(length(xnames))
  for (i in seq_along(xnames)) {
    xn <- xnames[i]
    xv <- df[[xn]]
    
    # Skip list/data-frame columns
    if (is.list(xv) && !is.atomic(xv)) {
      message(sprintf("Skipping '%s': predictor is a list/data-frame column.", xn))
      keep[i] <- FALSE
      next
    }
    
    # Optional coercion: "0"/"1" -> numeric
    if (coerce_binary_chars && is.character(xv)) {
      vals <- unique(na.omit(xv))
      if (length(vals) > 0 && all(vals %in% c("0", "1"))) {
        df[[xn]] <- as.numeric(xv)
        xv <- df[[xn]]
      }
    }
    
    # Skip non-variant predictors (all NA or <2 unique non-NA values)
    if (all(is.na(xv)) || length(unique(na.omit(xv))) < 2) {
      message(sprintf("Skipping '%s': predictor has <2 unique non-NA values.", xn))
      keep[i] <- FALSE
      next
    }
    
    keep[i] <- TRUE
  }
  xnames <- xnames[keep]
  
  if (length(xnames) == 0) {
    stop("No valid predictors found in the selected range after filtering.")
  }
  
  # --- Optional standardization of numeric predictors ---
  if (standardize) {
    for (xn in xnames) {
      if (is.numeric(df[[xn]])) {
        df[[xn]] <- as.numeric(scale(df[[xn]]))
      }
    }
    message("Numeric predictors standardized (mean=0, sd=1).")
  }
  
  # --- Build safe formula with backticks ---
  backtick <- function(nm) paste0("`", nm, "`")
  rhs <- paste(vapply(xnames, backtick, ""), collapse = " + ")
  f <- as.formula(paste(backtick(y), "~", rhs))
  
  # --- Fit model ---
  fit <- lm(f, data = df, na.action = na_action)
  
  # --- Print concise diagnostics ---
  cat("\n", strrep("=", 72), "\n", sep = "")
  cat("Multivariable Linear Model\n")
  cat(sprintf("Formula: %s\n", deparse(f)))
  cat(strrep("-", 72), "\n", sep = "")
  s <- summary(fit)
  print(s)
  
  # Extra diagnostics
  cat(strrep("-", 72), "\n", sep = "")
  cat(sprintf("Observations used: %d\n", s$df[1] + s$df[2]))
  cat(sprintf("Residual df: %d\n", s$df[2]))
  cat(sprintf("Residual SE: %.3f\n", s$sigma))
  cat(sprintf("Multiple R-squared: %.5f\n", s$r.squared))
  cat(sprintf("Adjusted R-squared: %.5f\n", s$adj.r.squared))
  suppressWarnings({
    aic_val <- try(AIC(fit), silent = TRUE)
    if (!inherits(aic_val, "try-error")) {
      cat(sprintf("AIC: %.3f\n", aic_val))
    }
  })
  
  invisible(fit)
}
```

### Running Multi LM Function

```{r}
mmodel <- run_multivar_lm(final_wide_trim)
```

### Building Backward Stepwise Regression Function

```{r}
run_stepwise_backward_lm <- function(df,
                                     y = "avg_salary_k",
                                     x_cols = 12:34,
                                     na_action = na.exclude,
                                     sanitize_names = FALSE,
                                     coerce_binary_chars = TRUE,
                                     standardize = FALSE,
                                     criterion = c("AIC", "BIC"),
                                     trace = TRUE) {
  criterion <- match.arg(criterion)
  
  # --- Basic checks ---
  if (!y %in% names(df)) {
    stop(sprintf("Response column '%s' not found in data.", y))
  }
  if (any(x_cols < 1 | x_cols > ncol(df))) {
    stop("x_cols index range is out of bounds for the provided data frame.")
  }
  
  # --- Optionally sanitize column names ---
  if (sanitize_names) {
    names(df) <- make.unique(make.names(names(df)))
    if (!y %in% names(df)) {
      stop("After sanitizing names, the response column was renamed; update 'y' to the new name.")
    }
    message("Column names were sanitized via make.names() + make.unique().")
  }
  
  # --- Ensure response is numeric ---
  if (!is.numeric(df[[y]])) {
    suppressWarnings({
      df[[y]] <- as.numeric(df[[y]])
    })
    if (anyNA(df[[y]])) {
      warning(sprintf("Coercing '%s' to numeric introduced NAs; check your response column.", y))
    }
  }
  
  # --- Collect candidate predictors ---
  xnames <- names(df)[x_cols]
  
  # --- Filter: skip problematic types, coerce binary chars, skip constants ---
  keep <- logical(length(xnames))
  for (i in seq_along(xnames)) {
    xn <- xnames[i]
    xv <- df[[xn]]
    
    # Skip list/data-frame columns
    if (is.list(xv) && !is.atomic(xv)) {
      message(sprintf("Skipping '%s': predictor is a list/data-frame column.", xn))
      keep[i] <- FALSE
      next
    }
    
    # Optional coercion: "0"/"1" -> numeric
    if (coerce_binary_chars && is.character(xv)) {
      vals <- unique(na.omit(xv))
      if (length(vals) > 0 && all(vals %in% c("0", "1"))) {
        df[[xn]] <- as.numeric(xv)
        xv <- df[[xn]]
      }
    }
    
    # Skip non-variant predictors (all NA or <2 unique non-NA values)
    if (all(is.na(xv)) || length(unique(na.omit(xv))) < 2) {
      message(sprintf("Skipping '%s': predictor has <2 unique non-NA values.", xn))
      keep[i] <- FALSE
      next
    }
    
    keep[i] <- TRUE
  }
  xnames <- xnames[keep]
  
  if (length(xnames) == 0) {
    stop("No valid predictors found in the selected range after filtering.")
  }
  
  # --- Optional standardization of numeric predictors ---
  if (standardize) {
    for (xn in xnames) {
      if (is.numeric(df[[xn]])) {
        df[[xn]] <- as.numeric(scale(df[[xn]]))
      }
    }
    message("Numeric predictors standardized (mean=0, sd=1).")
  }
  
  # --- Build safe formula with backticks ---
  backtick <- function(nm) paste0("`", nm, "`")
  rhs <- paste(vapply(xnames, backtick, ""), collapse = " + ")
  f_full <- as.formula(paste(backtick(y), "~", rhs))
  
  # --- Fit full model ---
  fit_full <- lm(f_full, data = df, na.action = na_action)
  
  # --- Choose penalty k (AIC vs BIC) ---
  n <- nobs(fit_full)  # number of observations actually used
  k <- if (criterion == "AIC") 2 else log(n)
  
  # --- Backward stepwise using stats::step ---
  fit_step <- step(fit_full,
                   direction = "backward",
                   trace = as.integer(trace),
                   k = k)
  
  # --- Print concise diagnostics for the final selected model ---
  cat("\n", strrep("=", 72), "\n", sep = "")
  cat(sprintf("Backward Stepwise Linear Model (criterion: %s, k = %.3f)\n", criterion, k))
  cat(sprintf("Final Formula: %s\n", deparse(formula(fit_step))))
  cat(strrep("-", 72), "\n", sep = "")
  
  s <- summary(fit_step)
  print(s)
  
  # Which predictors were selected?
  selected <- attr(terms(fit_step), "term.labels")
  dropped  <- setdiff(xnames, selected)
  
  cat(strrep("-", 72), "\n", sep = "")
  cat("Selected predictors:\n")
  if (length(selected) > 0) {
    cat(paste0(" - ", selected), sep = "\n")
  } else {
    cat(" (none; intercept-only model)\n")
  }
  
  cat("\nDropped predictors:\n")
  if (length(dropped) > 0) {
    cat(paste0(" - ", dropped), sep = "\n")
  } else {
    cat(" (none dropped)\n")
  }
  
  # Extra metrics
  cat("\n", strrep("-", 72), "\n", sep = "")
  cat(sprintf("Observations used: %d\n", s$df[1] + s$df[2]))
  cat(sprintf("Residual df: %d\n", s$df[2]))
  cat(sprintf("Residual SE: %.3f\n", s$sigma))
  cat(sprintf("Multiple R-squared: %.5f\n", s$r.squared))
  cat(sprintf("Adjusted R-squared: %.5f\n", s$adj.r.squared))
  cat(sprintf("%s: %.3f\n", criterion, AIC(fit_step) + (k - 2) * length(coef(fit_step))))
  # Note: AIC() returns AIC with k = 2. For BIC, we report adjusted by (k-2)*p.
  
  invisible(fit_step)
}
```

### Running Backward Stepwise Regression Function

```{r}
stepmodel <- run_stepwise_backward_lm(final_wide_trim)
```

Google Gemini. (2026). Gemini 3 Flash \[Large language model\]. <https://gemini.google.com.> Accessed March 20, 2026.
